---
title: "dec01.Rmd"
author: "Riley King"
date: "12/4/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  eval = TRUE,
  message = FALSE,
  warning = FALSE,
  dpi = 300,
  fig.align = "center",
  fig.width = 11,
  fig.height = 7
)
```

```{r include=FALSE}
library(tidyverse)
library(here)
library(slider)
library(vapoRwave)
library(patchwork)
```

#Part 1

## Read in data
```{r}
raw_tbl <- read_lines(here("dec01/input.txt")) %>%
  as_tibble()

clean_tbl <- raw_tbl %>%
  mutate(value = as.double(value))
```

## Find differences between rows and determine if the change is an increase

```{r}
inc_dec_tbl <- clean_tbl %>%
  mutate(diff = value - lag(value, n = 1)) %>%
  mutate(inc_dec = case_when(
    diff > 0 ~ "(increased)",
    diff < 0 ~ "(decreased)",
    diff == 0 ~ "(no_change)",
    TRUE ~ "N/A - no previous measurement"
  ))

inc_dec_tbl
```

## Count Instances of Increase
```{r}
inc_dec_tbl %>%
  count(inc_dec)
```

There are __n=1462__ cases of increasing steps in incremental steps.

# Part 2 

## Calculate rolling windown using **slider** package

Use slide_dbl() function to calculate sum within rolling window of 3 and reclaculate diffs for each step 

```{r}
roll_3_tbl <- inc_dec_tbl %>%
  mutate(value_r3 = slide_dbl(.x = value, .f = sum, .before = 2, .complete = TRUE)) %>%
  mutate(diffs_r3 = value_r3 - lag(value_r3, n = 1)) %>%
  mutate(inc_dec_r3 = case_when(
    diffs_r3 > 0 ~ "(increased)",
    diffs_r3 < 0 ~ "(decreased)",
    diffs_r3 == 0 ~ "(no_change)",
    TRUE ~ "N/A - no previous measurement"
  ))

roll_3_tbl
```

## Count instances of increasing incremental change in rolling window (looking back 2 each time)

```{r}
roll_3_tbl %>%
  count(inc_dec_r3)
```

# Learnings

There are __n=1497__ cases of increasing steps in incremental rolling windows with a look-back period of 2.

>       * There are other packages that will calculate functions over a rolling window including Tidyquant which I'm fond of but slider would let you roll over dates which the other packages can't - that's the reason I would choose to use it going forward even though it wasn't needed here.

>       * Remember that if you want a rolling window of 3 measurements (as requested in this challenge) then the .before argument is set to 2 indicating a lookback period of 2 *previous* values.  Don't be tempted to state 3;  it's the lookback number not the total number.

>       * .complete = TRUE forces the slide() function to only display the rolling outputs involving a complete set of n=3 input values (as requested in this challenge).  Omitting this gets you some partial sums and will count to an incorrect value.


# Visualize (for fun)

## Plot Depth vs. Scan Increment
```{r}
depth_plt <- roll_3_tbl %>%
  mutate(row_id = row_number()) %>%
  ggplot(aes(x = row_id, y = value)) +
  geom_line(color = "limegreen") +
  labs(
    title = "Sonar Readings of Seafloor - Advent of Code 2021, Dec01",
    subtitle = "Let's find those mf keys!!!",
    x = "",
    y = "Depth"
  ) +
  scale_y_reverse() +
  new_retro()

depth_plt
```


## Plot Change in Depth for Each Step and then Rolling Window of 3
```{r}
delta_plt <- roll_3_tbl %>%
  mutate(row_id = row_number()) %>%
  rename(
    incremental_change = diff,
    rolling_window_3 = diffs_r3
  ) %>%
  pivot_longer(cols = c(incremental_change, rolling_window_3), names_to = "type", values_to = "val") %>%
  ggplot(aes(x = row_id, y = val)) +
  geom_line(aes(color = type), alpha = .8) +
  labs(
    x = "Scan Distance from Sub (Sonar Steps)",
    y = "Change in\n Depth"
  ) +
  new_retro() +
  theme(legend.title = element_blank())

delta_plt
```


## Combine with Patchwork

```{r}
(depth_plt/delta_plt) +
  plot_layout(guides = 'collect')
```

