---
title: "advent_of_code_2021_dec03"
author: "Riley King"
date: "12/6/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  eval = TRUE,
  message = FALSE,
  warning = FALSE,
  dpi = 300,
  fig.align = "center",
  fig.width = 11,
  fig.height = 7
)
```

```{r include=FALSE}
library(tidyverse)
library(here)
library(slider)
library(vapoRwave)
library(patchwork)
library(ggforce)
```

#Part 1

## Read in data

```{r}

raw_tbl <- read_lines(here("dec03/input.txt")) %>%
  as_tibble()
```

## Split the Strings

```{r}
bits_tbl <- raw_tbl$value %>%
  str_split("", simplify = TRUE) %>%
  as_tibble()
```

## Reshape, Nest, Count

```{r}

bit_count_tbl <- bits_tbl %>%
  pivot_longer(cols = everything()) %>%
  mutate(
    name = as_factor(name),
    value = as.integer(value)
  ) %>%
  group_by(name) %>%
  summarize(bit_sums = sum(value)) %>%
  mutate(mcb = case_when(
    bit_sums >= nrow(bits_tbl) / 2 ~ 1,
    bit_sums < 500 ~ 0
  )) %>%
  ungroup() %>%
  mutate(inverse_mcb = case_when(
    mcb == 1 ~ 0,
    TRUE ~ 1
  ))

bit_count_tbl
```

## Reshape and Assemble Gamma and Epsilon

```{r}
gamma <- bit_count_tbl %>%
  select(name, mcb) %>%
  pivot_wider(id_cols = everything(), names_from = "name", values_from = "mcb") %>%
  unite("binary", V1:V12, sep = "")

epsilon <- bit_count_tbl %>%
  select(name, inverse_mcb) %>%
  pivot_wider(id_cols = everything(), names_from = "name", values_from = "inverse_mcb") %>%
  unite("binary", V1:V12, sep = "")
```

## Convert from Binary to Decimal and Multiply for Results
```{r}
summary_tbl <- bind_rows(gamma, epsilon) %>%
  mutate(param = c("gamma", "epsilon")) %>%
  select(param, everything()) %>%
  mutate(decimal = strtoi(binary, base = 2))

summary_tbl %>%
  summarize(answer = prod(decimal))
```

The answer to Part 1 is 3549854

# Part 2

## Find Oxygen Generator Rating

Going to try a while loop.  

```{r}
# Initialize Values
start_tbl <- raw_tbl
i_count <- 1
row_counter <- nrow(start_tbl)

while (row_counter > 1) {
  temp_tbl <- start_tbl$value %>%
    str_split("", simplify = TRUE) %>%
    as_tibble() %>%
    bind_cols(start_tbl) %>%
    select(value, everything()) %>%
    select(1, 1 + i_count) %>%
    mutate_at(.vars = 2, as.numeric) %>%
    rename(new_var = 2)

  start_tbl <- temp_tbl %>%
    mutate(more = case_when(
      mean(temp_tbl[[2]]) >= .5 ~ 1,
      TRUE ~ 0
    )) %>%
    filter(more == new_var) %>%
    select(value)

  i_count <- i_count + 1

  row_counter <- nrow(start_tbl)
}

oxygen_tbl <- start_tbl %>%
  mutate(decimal = strtoi(value, base = 2))

oxygen_tbl
```

## Find CO2 Scrubber Rating

```{r}
# Initialize Values
start_tbl <- raw_tbl
i_count <- 1
row_counter <- nrow(start_tbl)

while (row_counter > 1) {
  temp_tbl <- start_tbl$value %>%
    str_split("", simplify = TRUE) %>%
    as_tibble() %>%
    bind_cols(start_tbl) %>%
    select(value, everything()) %>%
    select(1, 1 + i_count) %>%
    mutate_at(.vars = 2, as.numeric) %>%
    rename(new_var = 2)

  start_tbl <- temp_tbl %>%
    mutate(more = case_when(
      mean(temp_tbl[[2]]) >= .5 ~ 1,
      TRUE ~ 0
    )) %>%
    mutate(less = case_when(
      more == 1 ~ 0,
      TRUE ~ 1
    )) %>%
    filter(less == new_var) %>%
    select(value)

  i_count <- i_count + 1

  row_counter <- nrow(start_tbl)
}

co2_tbl <- start_tbl %>%
  mutate(decimal = strtoi(value, base = 2))

co2_tbl
```

## Life Support Rating

```{r}
lsr <- oxygen_tbl[[1, 2]] * co2_tbl[[1, 2]]

lsr
```

The answer is: 3765399

# Learnings

So many! 

> *unite() does the opposite of separate; format is: unite("binary", V1:V12, sep = "")
> *convert binary to decimal using strtoi(): mutate(decimal = strtoi(value, base = 2))
> *i have never actually used a while loop outside of a toy problem and it was difficult for me to remember how; but I did!
> *i thought it was clever to use mean < 0.5 to determine if there were more 0's or 1's and then filter off of that value

# Visualize (for fun)

```{r }
bit_name_tbl <- tibble(
  x = "bit_",
  y = seq(1, 12, length.out = 12)
) %>%
  unite("new_name", x:y, sep = "")

bit_count_tbl %>%
  bind_cols(bit_name_tbl) %>%
  mutate(
    mcb = as_factor(mcb),
    new_name = as_factor(new_name)
  ) %>%
  # mutate(mcb = case_when(mcb == 1))
  ggplot(aes(x = new_name, y = bit_sums)) +
  geom_col(aes(fill = mcb)) +
  new_retro() +
  labs(
    x = "",
    y = "Sum of 1's",
    fill = "Most Common\n Bit in\n Position",
    title = "Sum of the 1's in each Bit Position",
    subtitle = "Bit sums > 500 indicate 1 is more common than 0"
  )
```
