---
title: "dec02.Rmd"
author: "Riley King"
date: "12/5/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  eval = TRUE,
  message = FALSE,
  warning = FALSE,
  dpi = 300,
  fig.align = "center",
  fig.width = 11,
  fig.height = 7
)
```

```{r include=FALSE}
library(tidyverse)
library(here)
library(slider)
library(vapoRwave)
library(patchwork)
library(ggforce)
```

#Part 1

## Read in data
```{r}

raw_tbl <- read_lines(here("dec02/input.txt")) %>%
  as_tibble()

```

## Separate Direction and Magnitude, Add cols for forward and depth steps

### Confirm there are no Direction = "Back" Steps (None were included in example)

```{r}
raw_tbl %>%
  separate(col = value, into = c("direction", "magnitude")) %>%
  count(direction)

```

As expected, there are no back steps. Proceed with determining steps of each type:

# Calculate Steps of Each Type

```{r}

steps_tbl <- raw_tbl %>%
  separate(col = value, into = c("direction", "magnitude")) %>%
  mutate(magnitude = as.double(magnitude)) %>%
  mutate(forward_step = case_when(direction == "forward" ~ magnitude,
                                  TRUE ~ 0)) %>%
  mutate(depth_step = case_when(direction == "down" ~ magnitude,
                                direction == "up" ~ magnitude * (-1),
                                TRUE ~ 0))
```

## Aggregate Sums of Steps and Multiply Final Answers

```{r}
step_sum_tbl <- steps_tbl %>% 
  summarize(forward_sum = sum(forward_step),
            depth_sum = sum(depth_step)) %>%
  mutate(step_prod = forward_sum * depth_sum) %>%
  ungroup()

answer <- step_sum_tbl %>% pluck("step_prod")

print(str_glue("The answer is {answer}"))
```

# Part 2

## Convert to new movement scheme

Fortunately we aleady had the list of incremental steps so we just rename depth step to aim step.  We need a col to track the cumuluative sum of the aim step which I call "current aim".  Then the depth step is the product of current aim * forward step
```{r}

aim_tbl <- steps_tbl %>%
  rename(aim_step = depth_step) %>%
  mutate(current_aim = cumsum(aim_step)) %>%
  mutate(depth_step = current_aim * forward_step)

aim_tbl
```

## Aggregate Horizontal Position and Depth Sums

```{r}
agg_tbl <- aim_tbl %>%
  summarize(horizontal_sum = sum(forward_step),
            depth_sum = sum(depth_step)) %>%
  mutate(prod = horizontal_sum * depth_sum)

agg_tbl
```

## Extract Answer

```{r}
answer_2 <- agg_tbl %>% pluck("prod")

print(str_glue("The answer is {answer_2}"))
```

# Visualize (for fun)

This was wrong but made a really cool accidental art image!

```{r}
steps_tbl %>%
  ggplot(aes(x = forward_step, y = depth_step)) +
  geom_path(color = "limegreen") +
  new_retro()
```
Do it right!


## Bind Results from Parts 1 and 2

```{r}
summary_tbl <- aim_tbl %>%
  mutate(depth_using_aim = cumsum(depth_step)) %>%
  select(depth_using_aim) %>%
  bind_cols(steps_tbl) %>%
  mutate(depth_without_aim = cumsum(depth_step),
         forward_position = cumsum(forward_step)) %>%
  select(direction, forward_position, depth_without_aim, depth_using_aim) 

summary_tbl
```

## Reshape and Plot

```{r}
summary_tbl %>%
  pivot_longer(cols = c(depth_without_aim, depth_using_aim), names_to = "depth_mode", values_to = "depth_value") %>%
  ggplot(aes(x = forward_position, y = depth_value)) +
  geom_path(aes(color = depth_mode)) +
  labs(
    x = "Forward Position",
    y = "Depth Position",
    title = "Horizontal and Depth Tracking of the Sub - Advent of Code 2021, Dec02",
    subtitle = "We must go deeper!"
  ) +
  scale_y_continuous(trans = trans_reverser('log10')) +
  new_retro() +
  theme(legend.title = element_blank())
```


# Learnings

> *Tidyverse rocks and made a hard problem pretty easy (at least the way my brain works)
> *If you want to reverse a log scale axis, the easiest way (after Googling) is the ggforce package which has:   scale_y_continuous(trans = trans_reverser('log10')) 

see: https://stackoverflow.com/questions/11053899/how-to-get-a-reversed-log10-scale-in-ggplot2
