---
title: "advent_of_code_2021_dec04.Rmd"
author: "Riley King"
date: "12/7/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  eval = TRUE,
  message = FALSE,
  warning = FALSE,
  dpi = 300,
  fig.align = "center",
  fig.width = 11,
  fig.height = 7
)
```

```{r include=FALSE}
library(tidyverse)
library(here)
library(slider)
library(vapoRwave)
library(patchwork)
```


# Read in Data

## Draws

```{r}
raw_draws_tbl <- read_lines(here("dec04/input.txt"), n_max = 1) %>%
  as_tibble()

draws_tbl <- tibble(draws = as.vector(str_split(raw_draws_tbl$value, pattern = ","))) %>% 
  unnest() %>%
  mutate(draws = as.double(draws))

draws_tbl 
```
## Boards

```{r}
boards_tbl <- tibble(square_value = as.integer(scan(here("dec04/input.txt"), what = "list", skip = 2))) %>%
  mutate(board_number = rep(1:(2500/25), each=25),
         col_number = rep(1:5, times = 2500/5),
         row_number = rep(rep(1:5, each = 5), times = 2500/25)) %>%
  mutate(match_flag = 0) 

boards_tbl
```

## Simulate the Game

```{r}

i_count <- 1
boards_loop_tbl <- boards_tbl
max_row_sum <- 0
max_col_sum <- 0
rc_sum <- 0

while(rc_sum < 5){

this_draw <- draws_tbl[[i_count,1]]

boards_loop_tbl <- boards_loop_tbl %>%
  mutate(match_flag = case_when(square_value == this_draw ~ 1,
                         TRUE ~ match_flag))

row_score_tbl <- boards_loop_tbl %>%
  group_by(board_number, row_number) %>%
  summarize(row_sum = sum(match_flag)) %>%
  ungroup()

col_score_tbl <- boards_loop_tbl %>%
  group_by(board_number, col_number) %>%
  summarize(col_sum = sum(match_flag)) %>%
  ungroup()

max_row_sum <- max(row_score_tbl$row_sum)
max_col_sum <- max(col_score_tbl$col_sum)

rc_sum <- max(max_row_sum, max_col_sum)

i_count <- i_count + 1

}
```


## Identify Winning Board

```{r}
col_score_tbl %>%
  filter(col_sum == 5) 
```

Board number 17 is the winner with a win down column 1

## Peek at Winning Board 17 

```{r}
boards_loop_tbl %>%
  filter(board_number == 17)
```

## Get Sum of Unmarked Numbers

```{r}
boards_loop_tbl %>%
  filter(board_number == 17) %>%
  group_by(match_flag) %>%
  summarize(sum_unmarked = sum(square_value))

```

The sum of the unmarked numbers is 858.

## Get Number that Was Just Called

```{r}

numbered_draws <- draws_tbl %>%
  mutate(draw_sequence = row_number()) %>%
  rename(square_value = draws)

boards_loop_tbl %>%
  filter(board_number == 17,
         match_flag == 1) %>%
  left_join(numbered_draws, by = "square_value") %>%
  arrange(desc(draw_sequence))

```

The number just called was 39

# Get Answer to Pt 1

```{r}
39 * 858
```

The answer to Part 1 is 33462

# Part 2

## Which Board Wins Last


```{r}

i_count <- 1
boards_loop_tbl <- boards_tbl
bingo_counter <-  0
all_bingo_tbl <- tibble(board_number = 1000,
                        max_row_sum = 1000,
                        max_col_sum = 1000,
                        max_either = 1000,
                        bingo_yet = 1000,
                        winning_draw = 1000)
```

```{r}

while(bingo_counter < 101){

this_draw <- draws_tbl[[i_count,1]]

boards_loop_tbl <- boards_loop_tbl %>%
  mutate(match_flag = case_when(square_value == this_draw ~ 1,
                         TRUE ~ match_flag))

row_score_tbl <- boards_loop_tbl %>%
  group_by(board_number, row_number) %>%
  summarize(row_sum = sum(match_flag)) %>%
  ungroup() %>%
  group_by(board_number) %>%
  summarize(max_row_sum = max(row_sum)) %>%
  ungroup()

col_score_tbl <- boards_loop_tbl %>%
  group_by(board_number, col_number) %>%
  summarize(col_sum = sum(match_flag)) %>%
  ungroup() %>%
  group_by(board_number) %>%
  summarize(max_col_sum = max(col_sum)) %>%
  ungroup()

remaining_boards_tbl <- row_score_tbl %>%
  inner_join(col_score_tbl) %>%
  rowwise() %>%
  mutate(max_either = max(max_row_sum, max_col_sum)) %>%
  ungroup() %>%
  mutate(bingo_yet= case_when(max_either == 5 ~ 1,
                              TRUE ~ 0)) %>%
  mutate(winning_draw = this_draw) %>%
  ungroup() 

this_bingo_tbl <- remaining_boards_tbl %>%
  filter(bingo_yet == 1)

all_bingo_tbl <- all_bingo_tbl %>% 
  bind_rows(this_bingo_tbl)

bingo_counter <- nrow(all_bingo_tbl)

if(bingo_counter != 101){
  
  boards_loop_tbl <- boards_loop_tbl %>%
  filter(board_number %notin% this_bingo_tbl$board_number)

}else{
    break
  }

i_count <- i_count + 1
  

}


```

## Get Sum of Unmarked Squares

```{r}
boards_loop_tbl %>%
  filter(match_flag == 0) %>%
  summarize(sum = sum(square_value))
```

The sum of the unmarked squares is 310

## Get Last Number Drawn
```{r}
all_bingo_tbl 
```

The winning draw was 97

## Get Answer to Part 2

```{r}
97*310
```

# Learnings

> *scan() function helped read in messy data
> *%notin% custom operator helped for filtering
> *if/then statements in a while loop with 'break' keyword to exit as needed
> *control flow in general 

# Visualize

```{r}
boards_loop_tbl
```

```{r}
boards_loop_tbl %>% 
  mutate(match_flag = as_factor(match_flag)) %>%
  ggplot(aes(x = col_number, y = row_number, fill = match_flag)) +
  geom_tile(aes(color = "black")) +
  geom_segment(aes(x = 1, y = 1, xend = 1, yend = 5), linetype = 2, color = "limegreen") +
geom_text(aes(label = square_value), color = "black", show.legend = FALSE) +
  labs(
    title = "Bingo vs. a Giant Squid - Advent of Code 2021, Dec04",
    subtitle = "Let Nature Win!!! - This is the Slowest Bingo Board",
    x = "",
    y = "",
    caption = "winning line shown in column 1"
  ) +
  new_retro()  + 
  scale_colour_newRetro() + 
  scale_fill_newRetro() +
  theme(legend.position = "none") 
  
```

